---
title: "Workflow"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This is a description of the workflow for *cytofQC*, which applies QC labels to each observation in a CyTOF dataset. The vignette shows how to use the general function \code{labelQC} to obtain labels and probabilities for all of the observations in a CyTOF data set. We then demonstrate the inner workings of the function to illustrate how it works for each observation type and show how to use the individual modeling functions that make up \code{labelQC}. we then end with a demonstration of how the labeled data cluster using UMAP. 


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  message = FALSE, 
  warning = FALSE
)
```

```{r setup}
library(cytofQC)
library(ggplot2)
```

# Read in data and create initial dataset

The function \code{dataPrep} read in the data and formats a \code{data.frame} that contains all of the QC variables. The variables are renamed so that their names can be easily accessed by downstream functions. The help file specifies how to do this. The Gaussian parameters and event length are transformed with the \code{log1p} function. The output is a \code{matrix}. 

```{r}
f <- "../../data/FlowRepository_FR-FCM-Z29V_files/REP_1_deid.fcs"
tech <- dataPrep(f)
head(tech)
```

# Use labelQC to obtain labels for each obsevation

The integral function of cytofQC is \code{labelQC}. A \code{data.frame} obtained from \code{dataPrep} can be passed to \code{labelQC} with no other arguments and it will return a \code{data.frame} that contains all of the labels and the probability that each observation belongs to each of the specified labels. The data we are using for this vignette has only live cells so we call \code{labelQC} in a way that omits the labeling of the live/dead cells. The following call will label the cells using a support vector machine and mutual nearest neighbors. 

```{r}
labels <- labelQC(tech, types = c("bead", "doublet", "debris"))
head(labels)
table(labels$label)
```


# How does labelQC work?

The following workflow demonstrates what is happening in the \code{labelQC} function. The user can use any of these functions to customize a labeling algorithm. We start with creating the basic labeling \code{data.frame}. The function \code{qcDataFrame} creates a \code{data.frame} that contains the labels and information about the labels. It is built using the \code{tech} dataset. The first variable is the label and each observation is assigned a label of either *cell* or *GDPzero*. The *GDPzero* label is assigned to any variable that has a 0 for any of the GDP variables. The \code{data.frame} has variables for beads, doublets, debris, and viability. The values for those variables are initially set to *NA*. The purpose of this dataset is to iteratively change the label for the observations and assign probabilities to each observation in regard its status as a bead, doublet, debris, or live/dead cell. 

```{r}
labels <- qcDataFrame(tech)
head(labels)
table(labels$label)
```


# Nearest neighbors

The next piece in the process is finding the nearest neighbors for each of the observations. There are currently two choices: k-means nearest neighbors and mutual nearest neighbors. This piece only needs to be done once. 

```{r eval = FALSE}
# K-means nearest neighbors
knn <- NN(tech, n = 10, type = "knn")
head(knn)
dim(knn)

```

```{r}
# Mutual nearest neighbors
mnn <- NN(tech, n = 20, type = "mnn")
head(mnn)
dim(mnn)

```

# Beads

## Initial bead labeling

The beads are typically the first type of observation that should be labeled. The beads should separate clearly from the non-beads. The first step is to obtain initial labels for the beads. This is done with the \code{initialBead} function. The function returns a \code{data.frame} that contains the bead assignment for each bead channel and a final bead designation in the variable called *init*. 

```{r}
beads <- initialBead(tech, labels)
head(beads)
sum(beads$init)
```

The function \code{cytofHist} can be used to visualize the data with regard to a grouping variable. This allows us to see what is classified as beads in the original bead data. TRUE indicates observations that are initially characterized as beads. 

```{r}
beadPlots <- NULL
for (i in 1:4) {
  beadPlots[[i]] <- cytofHist(x = tech[, i + 1], group = beads$init, 
                             title = paste("Bead channel", i))
}

```
```{r}
ggpubr::ggarrange(plotlist = beadPlots, ncol = 2, nrow = 2)
```
The zeros make it difficult to see the trends, so here are the plots without the zeros. The density option can be used to emphasize the smaller group. 

```{r}
beadPlots2 <- NULL
for (i in 1:4) {
  tmp <- data.frame(bead = tech[, i + 1], group = beads$init)
  tmp <- tmp[tmp$bead > 0, ]
  beadPlots2[[i]] <- cytofHist(x = tmp$bead, group = tmp$group, 
                               type = "density", 
                               title = paste("Bead channel", i))
}

```
```{r}
ggpubr::ggarrange(plotlist = beadPlots2, ncol = 2, nrow = 2)
```

## Identify Matches

This next part identifies the observations that are matched with all of its neighbors in regard to bead assignment using the nearest neighbors or mutual nearest neighbors matrix. If an observation is a designated as a bead, its nearest neighbors that are also beads are assigned a value of FALSE. If a neighbor is not a bead, it is given a value of TRUE, indicating that it is a mismatch of the observation. The rows are then summed together and rows with no TRUE values are designated as matches. The function returns a matrix with values of TRUE/FALSE as values indicating the mismatch status and a variable called *match* that is TRUE if the observation is a mismatch and FALSE if it is a match. 

The user can change how strict the threshold for being a match is by adjusting the *threshold* argument. The default is that all of the neighbors must be a match for the observation to be given a value of TRUE for the *match* variable.  


```{r}
sure <- surematch(mnn, beads$init, threshold = 0)
head(sure)
sum(sure$Mismatch)
```

## Create a subset of data to train a model

We plan to create a model using a smaller set of data. This next piece selects a subset of data points to build the model. Only the matches selected by the previous piece are considered to be in this subset. If no points are designated as sure matches, the entire set of unclassified data points (those that still have a *cell* label) is sampled. The data are sampled at random, but weighted to ensure we have a fairly balanced dataset to build the labeling model with. The function \code{modelData|} returns a set of indices that identifies the points in the dataset that will be used to compute the labeling model. 

```{r}
ind <- modelData(labels, subset = sure$match, init = beads$init, n = 4000)
length(ind)
head(ind)

```

The following histogram shows where the selected points are relative to the rest of the points for each of the bead channels. Note that TRUE designates the points that were selected to train the model. 

```{r}
index <- rep(FALSE, nrow(tech))
index[ind] <- TRUE
beadPlots3 <- NULL
for (i in 1:4) {
  beadPlots3[[i]] <- cytofHist(x = tech[, i + 1], group = index, 
                               type = "density", 
                               title = paste("Bead channel", i))
}

```

```{r}
ggpubr::ggarrange(plotlist = beadPlots3, ncol = 2, nrow = 2)
```

## Final classification of beads

There are currently three models that can be used to classify the beads: support vector machines (SVMs), random forests, and gradient boosting machines (GBMs). The method below uses random forests, but the process is identical for SVMs and GBMs. The functions for SVMs and GBMs are \code{svmLabel} and \code{gbmLabel}. The labeling functions take the same input and produce identical output. The output is a \code{data.frame} that is identical to the \code{qcDataFrame} but with the beads labeled as beads and the predicted probability of being a bead assigned to each observation in the *bead* variable. This is the *labels* \code{data.frame} that should be used for the next step because the *bead* label is needed for all other QC labels. 

```{r}
rf.labs <- rfLabel(tech, labels, type = "bead", init = beads$init, index = ind)
head(rf.labs)
table(rf.labs$label)

```

The following plots show a histogram of the bead measurement with the groups for each cell type shown in the plot. Note that the observations with a bead measurement of zero have been omitted from the plot because they dominate the histogram and make it difficult to see the three groups. 

```{r}
beadPlots <- NULL
for (i in 1:4) {
  tmp <- data.frame(bead = tech[, i + 1], group = rf.labs$label)
  tmp <- tmp[tmp$bead > 0, ]
  beadPlots[[i]] <- cytofHist(tmp$bead, group = tmp$group, type = "count", 
                              title = paste("Bead channel", i))
}

```

```{r}
ggpubr::ggarrange(plotlist = beadPlots, ncol = 2, nrow = 2)

```


# Doublets

This section shows how to label the doublets. It has the exact same structure as the beads. Note that the updated label \code{data.frame} should be used because the bead labels are needed for each of the steps. 


## Initial doublet labeling

The doublets are difficult to classify and we compute a doublet score to classify them. The method used to compute the doublet score is defined in the help file for the \code{initialDoublet} function. There are currently three options for classifying the doublet. They are selected using the score argument and they are documented in the help file for \code{initialDoublet}. The function returns the doublet score, the group it was assigned to, and the initial doublet classification in the *init* variable.   

```{r}
doublets <- initialDoublet(tech, rf.labs, score = 1)
head(doublets)

```

This histogram shows the initical classification of the doublet scores for the unclassified points (all of those not labeled *GDPzero* or *bead*). The scores for the classified points can be seen on the histogram by selection \code{na.rm = FALSE}. 

```{r}
cytofHist(doublets$doubletScore, doublets$doubletGroup, na.rm = TRUE, 
          title = "Doublet score and initial classification")
```



## Find matches

As with the beads, the next step is to identify the observations that are similar to each other in regard to the doublet score. The process is identical to the one used to assign the beads.  

```{r}
sureDoublet <- surematch(mnn, doublets$init, threshold = 0)
head(sureDoublet)
sum(sureDoublet$match)
```

The hisograms below show where the matched points are in relation to the rest of the points. This lets us see that the matched doublets are not too near the boundary. 

```{r}
mmPlots <- NULL
mmPlots[[1]] <- cytofHist(doublets$doubletScore, sureDoublet$match, type = "count", 
          title = "Matched doublets")
mmPlots[[2]] <- cytofHist(doublets$doubletScore, sureDoublet$match, type = "density", 
          title = "Matched doublets")

```
```{r fig.height = 3}
ggpubr::ggarrange(plotlist = mmPlots, ncol = 2, nrow = 1)
```


## Create a subset of data to train a doublet model

This code chunk shows how to select a random set of doublets and non-doublets to use to train a model.  

```{r}
d.rf.index <- modelData(rf.labs, subset = sureDoublet$match, init = doublets$init, n = 4000)

```

The following histogram shows where the selected points are relative to the rest of the doublet scores. 

```{r}
index <- rep("All", nrow(tech))
index[d.rf.index] <- "Subset"
cytofHist(x = doublets$doubletScore, group = index, type = "count", 
          title = "Doublet subset")

```

## Final classification of doublets

This shows the classification of doublets with random forests. Note that the label dataset from the last round of classifying is used as an input and the results are assigned to the same object. 

```{r}
rf.labs <- rfLabel(tech, rf.labs, type = "doublet", init = doublets$init, index = d.rf.index)
head(rf.labs)
table(rf.labs$label)

```

The following histogram shows doublet scores for the different groups that have been classified up to this point.

```{r}
cytofHist(doublets$doubletScore, rf.labs$label, type = "count", 
          title = "Labels on doublet score")
```

# Debris

This section shows how to label the debris. It has the exact same structure as the beads and doublets. Note that the updated label \code{data.frame} should be used because the previous labels are needed for this step. 


## Initial debris labeling

We compute a debris score to do a prelimnary classification of debris and non-debris. There are currently three options for classifying debris. They are selected using the score argument and they are documented in the help file for \code{initialDebris}. The function returns the debris score, the group it was assigned to, and the initial debris classification in the *init* variable.   

```{r}
debris <- initialDebris(tech, rf.labs, score = 1)
head(debris)

```
This histogram shows a classification of the debris scores for the unclassified points (all of those not labeled as *bead*, *doublet*, or *GDPzero*). The scores for the classified points can be seen on the histogram by selecting \code{na.rm = FALSE}. A label of 1 indicates a low debris score, or less likely to be debris, and a label of 2 indicates it is likely debris. 

```{r}
cytofHist(debris$debrisScore, debris$debrisGroup, na.rm = TRUE, 
          title = "Initial debris classification")
```

## Find mismatches

This next part identifies the observations that are mismatched from its neighbors in regard to debris assignment using the nearest neighbors matrix. 

```{r}
debm <- surematch(mnn, debris$init, threshold = 0)
head(debm)
sum(debm$Mismatch)
```

The following histogram shows the debris scores for the observations that are candidates for model training. Note that the matched observations are away from the boundary between the two groups. 

```{r}
mmPlots <- NULL
mmPlots[[1]] <- cytofHist(debris$debrisScore, debm$match, type = "count", 
          title = "Mismatched debris")
mmPlots[[2]] <- cytofHist(debris$debrisScore, debm$match, type = "density", 
          title = "Mismatched debris")

```
```{r fig.height = 3}
ggpubr::ggarrange(plotlist = mmPlots, ncol = 2, nrow = 1)
```

## Create a subset of data to train a debris model

The debris have enough matches to draw from to select a dataset from the matches. The following code chunk selects a training dataset for the model. 

```{r}
de.rf.index <- modelData(rf.labs, subset = debm$match, init = debris$init, n = 4000)

```

The following histogram shows where the selected points are relative to the rest of the debris scores. Note that the 'density' option is used so that the groups can be seen. When the 'count' option is used it is almost impossible to see the selected points on the histogram. 

```{r}
index <- rep("All", nrow(tech))
index[de.rf.index] <- "Subset"
cytofHist(x = debris$debrisScore, group = index, type = "density", 
          title = "Debris subset")

```

## Final classification of debris

The following code show how to do the final classification of debris with random forests. Note that the label dataset from the last round of classifying is used as an input and the results are assigned to the same object. 

```{r}
rf.labs <- rfLabel(tech, rf.labs, type = "debris", init = debris$init, index = de.rf.index)
head(rf.labs)
table(rf.labs$label)

```

This histogram shows the debris score for the observations that have been labeled up to this point. You can see that the observations that are labeled *debris* and *GDPzero*.  

```{r}
cytofHist(debris$debrisScore, rf.labs$label, type = "count", 
          title = "Labels on debris score")
```

# Final summary

The following shows the final labels on the different metrics used to classify the observations. 

```{r}
cytofHist(rowSums(tech[, c("Bead1", "Bead2", "Bead3", "Bead4")]), rf.labs$label, 
          title = "Final labels on Bead 4 measures")
tmp <- data.frame(bead = rowSums(tech[, c("Bead1", "Bead2", "Bead3", "Bead4")]), 
                  group = rf.labs$label)
tmp <- tmp[tmp$bead > 0, ]
cytofHist(tmp$bead, tmp$group, title = "Final labels on Bead measure - zeros removed")
cytofHist(doublets$doubletScore, rf.labs$label, title = "Final labels on doublet score")
cytofHist(debris$debrisScore, rf.labs$label, title = "Final labels on debris score")
```

This is a umap of the different labels. the umap was created using the standardized *tech* data. 

```{r}
s.tech <- scale(tech[, -1])
lab.umap <- uwot::umap(s.tech, ret_model = TRUE)
lab.umapD <- data.frame(x = lab.umap$embedding[, 1], y = lab.umap$embedding[, 2])

```

```{r}
umapData <- data.frame(lab.umapD, labels = rf.labs$label)
umapData$bead <- ifelse(rf.labs$label == "bead", "Bead", "Not-Bead")
umapData$doublet <- ifelse(rf.labs$label == "doublet", "Doublet", "Not-Doublet")
umapData$debris <- ifelse(rf.labs$label == "debris", "Debris", "Not-Debris")
umapData$GDPzero <- ifelse(rf.labs$label == "GDPzero", "GDPzero", "Not-GDPzero")
umapData$cell <- ifelse(rf.labs$label == "cell", "Cells", "Not-Cells")

umapData$labels <- factor(umapData$labels, levels = c("cell", "GDPzero", "bead", "doublet", "debris"))

```

```{r}
require(RColorBrewer)
cols <- brewer.pal(5, "Set2")
umapG <- ggplot(umapData, aes(x = x, y = y, color = labels)) + 
  geom_point(size = 0.25) + 
  scale_color_manual(values = cols) +
  labs(title = "Random forest", x = "", y = "") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  theme_bw()

umapG
```

Split out by cell type

```{r}
umaps <- NULL
cell_type <- levels(umapData$labels)

for (i in 1:5) {
  tmp <- data.frame(umapData[, 1:2], lab = umapData[, cell_type[i]])
  umaps[[i]] <- ggplot(tmp, aes(x = x, y = y, color = lab)) +
    geom_point(size = 0.25, show.legend = FALSE) +
    scale_color_manual("", values = c(cols[i], "gray80")) +
    labs(title = cell_type[i], x = "", y = "") +
    theme_bw()
}

```

```{r fig.height = 9}
library(gridExtra)
grid.arrange(umapG + guides(col = "none"), umaps[[1]], umaps[[2]], 
             umaps[[3]], umaps[[4]], umaps[[5]], nrow = 3, ncol = 2)
```


